---
title: "Predicting 30-Day Hospital Readmission for Patients with Diabetes"
output: html_document
date: "2026-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Preprocessing
Step 1:Load packages and read data
```{r}
library(tidyverse)
library(caret)

# Read the main dataset
diabetes <- read.csv("../data/diabetic_data.csv", stringsAsFactors = FALSE)

# Read the mapping file
ids_map <- read.csv("../data/IDS_mapping.csv", stringsAsFactors = FALSE)

# Check data size
dim(diabetes)

# Preview data
head(diabetes)
```
```{r}
names(diabetes)
```
Step 2: Replace “?” with missing values
```{r}
# Replace "?" with NA
diabetes[diabetes == "?"] <- NA

# Check missing values by variable
missing_summary <- data.frame(
  variable = names(diabetes),
  missing_count = colSums(is.na(diabetes)),
  missing_percent = colSums(is.na(diabetes)) / nrow(diabetes) * 100
)

# Show variables with missing values
missing_summary %>%
  filter(missing_count > 0) %>%
  arrange(desc(missing_percent))
```

Remove variables with too much missingness
```{r}
diabetes_clean <- diabetes %>%
  select(
    -weight,
    -medical_specialty,
    -payer_code
  )

# For the remaining categorical variables, replace missing values with "Unknown"
diabetes_clean <- diabetes_clean %>%
  mutate(across(where(is.character), ~replace_na(.x, "Unknown")))
```

Step 4: Create binary outcome variable
```{r}
# Create binary outcome variable
# Yes = readmitted within 30 days
# No  = not readmitted within 30 days or readmitted after 30 days

diabetes_clean <- diabetes_clean %>%
  mutate(
    readmitted_30 = ifelse(readmitted == "<30", "Yes", "No"),
    readmitted_30 = as.factor(readmitted_30)
  )

# Check the distribution of the new outcome variable
table(diabetes_clean$readmitted_30)
prop.table(table(diabetes_clean$readmitted_30))
```
Step 5: Remove ID variables and original outcome variable
```{r}
# Remove ID variables and the original outcome variable
# encounter_id and patient_nbr are identifiers, not useful predictors.
# readmitted is removed because we already created readmitted_30.

diabetes_clean <- diabetes_clean %>%
  select(
    -encounter_id,
    -patient_nbr,
    -readmitted
  )
```

Step 6: Convert ID-coded variables to factors
```{r}
# These variables are numerical codes, but they represent categories.
# Therefore, we convert them to factors.

diabetes_clean <- diabetes_clean %>%
  mutate(
    admission_type_id = as.factor(admission_type_id),
    discharge_disposition_id = as.factor(discharge_disposition_id),
    admission_source_id = as.factor(admission_source_id)
  )
```

Step 7: Convert all character variables to factors
```{r}
# Convert all remaining character variables to factors
# This is needed for classification models and logistic regression.

diabetes_clean <- diabetes_clean %>%
  mutate(across(where(is.character), as.factor))
```


Step 8: Group diagnosis codes
```{r}
# Function to group ICD diagnosis codes into broader disease categories

group_diag <- function(x) {
  
  x <- as.character(x)
  x[is.na(x)] <- "Unknown"
  
  # Some diagnosis codes start with letters, such as E or V.
  # We keep them as Other.
  x_num <- suppressWarnings(as.numeric(x))
  
  result <- case_when(
    x == "Unknown" ~ "Unknown",
    
    # Diabetes
    str_detect(x, "^250") ~ "Diabetes",
    
    # Circulatory system
    !is.na(x_num) & x_num >= 390 & x_num <= 459 ~ "Circulatory",
    x == "785" ~ "Circulatory",
    
    # Respiratory system
    !is.na(x_num) & x_num >= 460 & x_num <= 519 ~ "Respiratory",
    x == "786" ~ "Respiratory",
    
    # Digestive system
    !is.na(x_num) & x_num >= 520 & x_num <= 579 ~ "Digestive",
    x == "787" ~ "Digestive",
    
    # Injury
    !is.na(x_num) & x_num >= 800 & x_num <= 999 ~ "Injury",
    
    # Musculoskeletal system
    !is.na(x_num) & x_num >= 710 & x_num <= 739 ~ "Musculoskeletal",
    
    # Genitourinary system
    !is.na(x_num) & x_num >= 580 & x_num <= 629 ~ "Genitourinary",
    x == "788" ~ "Genitourinary",
    
    TRUE ~ "Other"
  )
  
  return(as.factor(result))
}

# Apply the grouping function to the three diagnosis variables

diabetes_clean <- diabetes_clean %>%
  mutate(
    diag_1_group = group_diag(diag_1),
    diag_2_group = group_diag(diag_2),
    diag_3_group = group_diag(diag_3)
  ) %>%
  select(
    -diag_1,
    -diag_2,
    -diag_3
  )

# Check the new diagnosis groups
table(diabetes_clean$diag_1_group)
table(diabetes_clean$diag_2_group)
table(diabetes_clean$diag_3_group)

```
Step 9: Check remaining missing values
```{r}
# Check if there are any remaining missing values

sum(is.na(diabetes_clean))

# Missing values by variable
colSums(is.na(diabetes_clean))[colSums(is.na(diabetes_clean)) > 0]
```

EDA

Outcome distribution
```{r}
# Outcome distribution
table(diabetes_clean$readmitted_30)

prop.table(table(diabetes_clean$readmitted_30))

library(ggplot2)

ggplot(diabetes_clean, aes(x = readmitted_30)) +
  geom_bar() +
  labs(
    title = "Distribution of 30-Day Readmission",
    x = "Readmitted within 30 days",
    y = "Count"
  )
```


EDA 2: Age distribution by readmission status
```{r}
ggplot(diabetes_clean, aes(x = age, fill = readmitted_30)) +
  geom_bar(position = "fill") +
  labs(
    title = "30-Day Readmission Rate by Age Group",
    x = "Age Group",
    y = "Proportion",
    fill = "Readmitted within 30 Days"
  )
```


EDA 3: Prior inpatient visits and readmission
```{r}
ggplot(diabetes_clean, aes(x = factor(number_inpatient), fill = readmitted_30)) +
  geom_bar(position = "fill") +
  labs(
    title = "30-Day Readmission Rate by Prior Inpatient Visits",
    x = "Number of Prior Inpatient Visits",
    y = "Proportion",
    fill = "Readmitted within 30 Days"
  )
```

EDA 4: Diagnosis group and readmission

```{r}
ggplot(diabetes_clean, aes(x = diag_1_group, fill = readmitted_30)) +
  geom_bar(position = "fill") +
  labs(
    title = "30-Day Readmission Rate by Primary Diagnosis Group",
    x = "Primary Diagnosis Group",
    y = "Proportion",
    fill = "Readmitted within 30 Days"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



Modeling

Train/test split
```{r}
library(caret)

# Make sure outcome is a factor
diabetes_clean$readmitted_30 <- factor(
  diabetes_clean$readmitted_30,
  levels = c("No", "Yes")
)

set.seed(123)

# 80/20 split
train_index <- createDataPartition(
  diabetes_clean$readmitted_30,
  p = 0.8,
  list = FALSE
)

train_data <- diabetes_clean[train_index, ]
test_data  <- diabetes_clean[-train_index, ]

# Drop unused factor levels
train_data <- droplevels(train_data)
test_data  <- droplevels(test_data)

# Check split size
dim(train_data)
dim(test_data)

# Check outcome distribution
prop.table(table(train_data$readmitted_30))
prop.table(table(test_data$readmitted_30))
```

```{r}
# Find factor variables with only one level in training data

one_level_vars <- names(train_data)[sapply(train_data, function(x) {
  is.factor(x) && nlevels(x) < 2
})]

one_level_vars

train_data_logit <- train_data %>%
  select(-all_of(one_level_vars))

test_data_logit <- test_data %>%
  select(-all_of(one_level_vars))

# Check again

names(train_data_logit)[sapply(train_data_logit, function(x) {
  is.factor(x) && nlevels(x) < 2
})]
```


```{r}
# Find factor variables that have levels in test data but not in training data

factor_vars <- names(train_data_logit)[sapply(train_data_logit, is.factor)]

new_level_vars <- c()

for (v in factor_vars) {
  train_levels <- levels(train_data_logit[[v]])
  test_levels  <- levels(test_data_logit[[v]])
  
  new_levels <- setdiff(test_levels, train_levels)
  
  if (length(new_levels) > 0) {
    cat("Variable:", v, "has new levels in test:", new_levels, "\n")
    new_level_vars <- c(new_level_vars, v)
  }
}

new_level_vars

# Remove variables that have unseen levels in the test set

train_data_logit <- train_data_logit %>%
  select(-all_of(new_level_vars))

test_data_logit <- test_data_logit %>%
  select(-all_of(new_level_vars))
```

Logistic Regression
```{r}
# Fit baseline logistic regression model

logit_model <- glm(
  readmitted_30 ~ .,
  data = train_data_logit,
  family = binomial
)

# Check model summary
summary(logit_model)
```


(a)Predict probabilities on test data
```{r}
# Predict probability of 30-day readmission on the test set

logit_prob <- predict(
  logit_model,
  newdata = test_data_logit,
  type = "response"
)

# Check first few predicted probabilities
head(logit_prob)
summary(logit_prob)
```
(b)confusion matrix
```{r}
library(caret)

# Convert probabilities to class labels using threshold = 0.5

logit_pred <- ifelse(logit_prob > 0.5, "Yes", "No")
logit_pred <- factor(logit_pred, levels = c("No", "Yes"))

# Confusion matrix

logit_cm <- confusionMatrix(
  data = logit_pred,
  reference = test_data_logit$readmitted_30,
  positive = "Yes"
)

logit_cm
```
```{r}
# Try threshold = 0.3

logit_pred_03 <- ifelse(logit_prob > 0.3, "Yes", "No")
logit_pred_03 <- factor(logit_pred_03, levels = c("No", "Yes"))

logit_cm_03 <- confusionMatrix(
  data = logit_pred_03,
  reference = test_data_logit$readmitted_30,
  positive = "Yes"
)

logit_cm_03
```
```{r}
# Try threshold = 0.2

logit_pred_02 <- ifelse(logit_prob > 0.2, "Yes", "No")
logit_pred_02 <- factor(logit_pred_02, levels = c("No", "Yes"))

logit_cm_02 <- confusionMatrix(
  data = logit_pred_02,
  reference = test_data_logit$readmitted_30,
  positive = "Yes"
)

logit_cm_02
```

(c)AUC and ROC curve
```{r}
library(pROC)

logit_roc <- roc(
  response = test_data_logit$readmitted_30,
  predictor = logit_prob,
  levels = c("No", "Yes")
)

logit_auc <- auc(logit_roc)
logit_auc

plot(
  logit_roc,
  main = "ROC Curve for Baseline Logistic Regression"
)
```

```{r}
threshold_results <- data.frame(
  Threshold = c(0.5, 0.3, 0.2),
  Accuracy = c(
    logit_cm$overall["Accuracy"],
    logit_cm_03$overall["Accuracy"],
    logit_cm_02$overall["Accuracy"]
  ),
  Sensitivity = c(
    logit_cm$byClass["Sensitivity"],
    logit_cm_03$byClass["Sensitivity"],
    logit_cm_02$byClass["Sensitivity"]
  ),
  Specificity = c(
    logit_cm$byClass["Specificity"],
    logit_cm_03$byClass["Specificity"],
    logit_cm_02$byClass["Specificity"]
  ),
  Balanced_Accuracy = c(
    logit_cm$byClass["Balanced Accuracy"],
    logit_cm_03$byClass["Balanced Accuracy"],
    logit_cm_02$byClass["Balanced Accuracy"]
  ),
  AUC = c(
    as.numeric(logit_auc),
    as.numeric(logit_auc),
    as.numeric(logit_auc)
  )
)

threshold_results
```

Classification Tree
```{r}
# Rename the cleaned train/test datasets to avoid confusion
# they have already been cleaned and can be reused for later models
train_model_data <- train_data_logit
test_model_data  <- test_data_logit
```

```{r}
library(tree)

# create classification tree
set.seed(123)

tree_model <- tree(
  readmitted_30 ~ .,
  data = train_model_data
)
```
```{r}
# plot tree
plot(tree_model)
text(tree_model, pretty = 0, cex = 0.5)
```

```{r}
# predict test set 
tree_prob <- predict(tree_model, newdata = test_model_data, type = "vector")[, "Yes"]

head(tree_prob)
summary(tree_prob)
```
```{r}
# try threshold = 0.5
tree_pred <- ifelse(tree_prob > 0.5, "Yes", "No")
tree_pred <- factor(tree_pred, levels = c("No", "Yes"))

tree_cm <- confusionMatrix(
  data = tree_pred,
  reference = test_model_data$readmitted_30,
  positive = "Yes"
)

tree_cm
```
```{r}
# try threshold = 0.3
tree_pred_03 <- ifelse(tree_prob > 0.3, "Yes", "No")
tree_pred_03 <- factor(tree_pred_03, levels = c("No", "Yes"))

tree_cm_03 <- confusionMatrix(
  data = tree_pred_03,
  reference = test_model_data$readmitted_30,
  positive = "Yes"
)

tree_cm_03
```
```{r}
# try threshold = 0.2
tree_pred_02 <- ifelse(tree_prob > 0.2, "Yes", "No")
tree_pred_02 <- factor(tree_pred_02, levels = c("No", "Yes"))

tree_cm_02 <- confusionMatrix(
  data = tree_pred_02,
  reference = test_model_data$readmitted_30,
  positive = "Yes"
)

tree_cm_02
```

```{r}
# ROC & AUC
tree_roc <- roc(
  response = test_model_data$readmitted_30,
  predictor = tree_prob,
  levels = c("No", "Yes")
)

tree_auc <- auc(tree_roc)
tree_auc

plot(tree_roc, main = "ROC Curve for Classification Tree")
```

```{r}
# results
tree_results <- data.frame(
  Threshold = c(0.5, 0.3, 0.2),
  Accuracy = c(
    tree_cm$overall["Accuracy"],
    tree_cm_03$overall["Accuracy"],
    tree_cm_02$overall["Accuracy"]
  ),
  Sensitivity = c(
    tree_cm$byClass["Sensitivity"],
    tree_cm_03$byClass["Sensitivity"],
    tree_cm_02$byClass["Sensitivity"]
  ),
  Specificity = c(
    tree_cm$byClass["Specificity"],
    tree_cm_03$byClass["Specificity"],
    tree_cm_02$byClass["Specificity"]
  ),
  Balanced_Accuracy = c(
    tree_cm$byClass["Balanced Accuracy"],
    tree_cm_03$byClass["Balanced Accuracy"],
    tree_cm_02$byClass["Balanced Accuracy"]
  ),
  AUC = c(
    as.numeric(tree_auc),
    as.numeric(tree_auc),
    as.numeric(tree_auc)
  )
)

tree_results
```

Random Forest
```{r}
library(randomForest)

train_model_data <- train_data_logit
test_model_data  <- test_data_logit

# Force factor levels in test data to match training data
for (v in names(train_model_data)) {
  if (is.factor(train_model_data[[v]])) {
    test_model_data[[v]] <- factor(
      test_model_data[[v]],
      levels = levels(train_model_data[[v]])
    )
  }
}

# create random forest
set.seed(123)

rf_model <- randomForest(
  readmitted_30 ~ .,
  data = train_model_data,
  ntree = 300,
  importance = TRUE
)

rf_model

# variable importance
importance(rf_model)
varImpPlot(rf_model, cex = 0.5)
```

```{r}
# predict prob on test set 
rf_prob <- predict(rf_model, newdata = test_model_data, type = "prob")[, "Yes"]

head(rf_prob)
summary(rf_prob)
```
```{r}
# try threshold = 0.5
rf_pred <- ifelse(rf_prob > 0.5, "Yes", "No")
rf_pred <- factor(rf_pred, levels = c("No", "Yes"))

rf_cm <- confusionMatrix(
  data = rf_pred,
  reference = test_model_data$readmitted_30,
  positive = "Yes"
)

rf_cm
```
```{r}
# try threshold = 0.3
rf_pred_03 <- ifelse(rf_prob > 0.3, "Yes", "No")
rf_pred_03 <- factor(rf_pred_03, levels = c("No", "Yes"))

rf_cm_03 <- confusionMatrix(
  data = rf_pred_03,
  reference = test_model_data$readmitted_30,
  positive = "Yes"
)

rf_cm_03
```
```{r}
# try threshold = 0.2
rf_pred_02 <- ifelse(rf_prob > 0.2, "Yes", "No")
rf_pred_02 <- factor(rf_pred_02, levels = c("No", "Yes"))

rf_cm_02 <- confusionMatrix(
  data = rf_pred_02,
  reference = test_model_data$readmitted_30,
  positive = "Yes"
)

rf_cm_02
```

```{r}
# ROC & AUC
rf_roc <- roc(
  response = test_model_data$readmitted_30,
  predictor = rf_prob,
  levels = c("No", "Yes")
)

rf_auc <- auc(rf_roc)
rf_auc

# plot
plot(rf_roc, main = "ROC Curve for Random Forest")
```

```{r}
# results
rf_results <- data.frame(
  Threshold = c(0.5, 0.3, 0.2),
  Accuracy = c(
    rf_cm$overall["Accuracy"],
    rf_cm_03$overall["Accuracy"],
    rf_cm_02$overall["Accuracy"]
  ),
  Sensitivity = c(
    rf_cm$byClass["Sensitivity"],
    rf_cm_03$byClass["Sensitivity"],
    rf_cm_02$byClass["Sensitivity"]
  ),
  Specificity = c(
    rf_cm$byClass["Specificity"],
    rf_cm_03$byClass["Specificity"],
    rf_cm_02$byClass["Specificity"]
  ),
  Balanced_Accuracy = c(
    rf_cm$byClass["Balanced Accuracy"],
    rf_cm_03$byClass["Balanced Accuracy"],
    rf_cm_02$byClass["Balanced Accuracy"]
  ),
  AUC = c(
    as.numeric(rf_auc),
    as.numeric(rf_auc),
    as.numeric(rf_auc)
  )
)

rf_results
```

Final Comparison
```{r}
model_com <- data.frame(
  Model = c("Logistic Regression", "Classification Tree", "Random Forest"),
  
  Accuracy_0.5 = c(
    logit_cm$overall["Accuracy"],
    tree_cm$overall["Accuracy"],
    rf_cm$overall["Accuracy"]
  ),
  Sensitivity_0.5 = c(
    logit_cm$byClass["Sensitivity"],
    tree_cm$byClass["Sensitivity"],
    rf_cm$byClass["Sensitivity"]
  ),
  Specificity_0.5 = c(
    logit_cm$byClass["Specificity"],
    tree_cm$byClass["Specificity"],
    rf_cm$byClass["Specificity"]
  ),
  Balanced_Accuracy_0.5 = c(
    logit_cm$byClass["Balanced Accuracy"],
    tree_cm$byClass["Balanced Accuracy"],
    rf_cm$byClass["Balanced Accuracy"]
  ),
  
  Accuracy_0.3 = c(
    logit_cm_03$overall["Accuracy"],
    tree_cm_03$overall["Accuracy"],
    rf_cm_03$overall["Accuracy"]
  ),
  Sensitivity_0.3 = c(
    logit_cm_03$byClass["Sensitivity"],
    tree_cm_03$byClass["Sensitivity"],
    rf_cm_03$byClass["Sensitivity"]
  ),
  Specificity_0.3 = c(
    logit_cm_03$byClass["Specificity"],
    tree_cm_03$byClass["Specificity"],
    rf_cm_03$byClass["Specificity"]
  ),
  Balanced_Accuracy_0.3 = c(
    logit_cm_03$byClass["Balanced Accuracy"],
    tree_cm_03$byClass["Balanced Accuracy"],
    rf_cm_03$byClass["Balanced Accuracy"]
  ),
  
  Accuracy_0.2 = c(
    logit_cm_02$overall["Accuracy"],
    tree_cm_02$overall["Accuracy"],
    rf_cm_02$overall["Accuracy"]
  ),
  Sensitivity_0.2 = c(
    logit_cm_02$byClass["Sensitivity"],
    tree_cm_02$byClass["Sensitivity"],
    rf_cm_02$byClass["Sensitivity"]
  ),
  Specificity_0.2 = c(
    logit_cm_02$byClass["Specificity"],
    tree_cm_02$byClass["Specificity"],
    rf_cm_02$byClass["Specificity"]
  ),
  Balanced_Accuracy_0.2 = c(
    logit_cm_02$byClass["Balanced Accuracy"],
    tree_cm_02$byClass["Balanced Accuracy"],
    rf_cm_02$byClass["Balanced Accuracy"]
  ),
  
  AUC = c(
    as.numeric(logit_auc),
    as.numeric(tree_auc),
    as.numeric(rf_auc)
  )
)

model_com
```
